---
title: "AMS532HW01"
author: "Khaliukova"
date: "8/28/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Q4 (a). Use the quilt.plot function to plot day 16 of the ozone data set and
add the US map on top of it. Following the example from help( quilt.plot)

```{r}
library(fields)
data( ozone2)
quilt.plot(ozone2$lon.lat, ozone2$y[16,])
US( add=TRUE, col="grey", lwd=2)
```

Q4. (b)
Now compare day 16 to both day 15 and day 17 in this data set. 

What are the actual dates for these cases? 

```{r}
summary(ozone2)
ozone2$dates
```
```{r}
attach(ozone2)
dates_actual<-as.Date(dates, "%y%m%d")

dates_actual[15]
dates_actual[16]
dates_actual[17]
```

Use the function set.panel( 1,3) to create a panel of 3 plots in a row

```{r}
set.panel(1,3)

quilt.plot( ozone2$lon.lat, ozone2$y[15,], main="Day 15", zlim= c( 0,165))
US( add=TRUE, col="grey", lwd=2)

quilt.plot( ozone2$lon.lat, ozone2$y[16,], main="Day 16", zlim= c( 0,165))
US( add=TRUE, col="grey", lwd=2)

quilt.plot( ozone2$lon.lat, ozone2$y[17,], main="Day 17", zlim= c( 0,165))
US( add=TRUE, col="grey", lwd=2)
```

What differences do you see between these days?
Ozone in the observed geographical region was at least 
50 parts per billion (ppb), and over 100 ppb in multiple locations on 
June 17th and 18th of 1987. However, on June 19th, 1987 much lower ozone 
levels were recorded. Ozone levels were between 10 and 65 ppb in the southern 
and  central parts of the region, northern part still observed the levels 
above 100 ppb. 

EXTRA CREDIT 1 See if you can find a better color scale than the default one 
(tim.colors) in these examples.

```{r}
library(viridis)
set.panel(1,3)
quilt.plot( ozone2$lon.lat, ozone2$y[15,], main="Day 15", zlim= c( 0,115))
US( add=TRUE, col=viridis(n = 33, alpha=0.25), lwd=2)

quilt.plot( ozone2$lon.lat, ozone2$y[16,], main="Day 16", zlim= c( 0,115))
US( add=TRUE, col=viridis(n = 33, alpha=0.25), lwd=2)

quilt.plot( ozone2$lon.lat, ozone2$y[17,], main="Day 17", zlim= c( 0,115))
US( add=TRUE, col=viridis(n = 33, alpha=0.25), lwd=2)
```


 EXTRA CREDIT 2 
- Where can one find the daily ozone monitoring data collected by local, state 
and US agencies? 
There are many agencies that monitor air quality: EPA, Centers for Disease 
Control and Prevention (CDC), NOAA National Weather Service, NASA Earth Science, 
National Park Service Air Resources, US Forest Service, National Association of 
Clean Air Agencies. 
The daily ozone monitoring data can be found on EPA website, go to Outdoor Air 
Quality Data page, select "Ozone" among Pollutants, years of interest, State, 
City, County, specify which monitor sites you would like to include. 
The records can be found on CDC websites. The ozone readings can be found on 
State Department of Public Health and Environment in Air Quality Section. 
  

What is the finest time scale on which ozone it is measured?
8-hour average ozone concentration is the finest time scale on which ozone 
is measured.

